---
title: "Network visualization with R"
output: html_notebook
---

http://kateto.net/network-visualization


```{r}
install.packages("igraph") 
install.packages("network") 
install.packages("sna")
install.packages("visNetwork")
install.packages("threejs")
install.packages("networkD3")
install.packages("ndtv")
```

# 2 Colors and fonts in R
## 2.1 Colors in R plots

```{r}
plot(x=1:10, y=rep(5,10), pch=19, cex=3, col="dark red")
points(x=1:10, y=rep(6, 10), pch=19, cex=3, col="557799")
points(x=1:10, y=rep(4, 10), pch=19, cex=3, col=rgb(.25, .5, .3))
```


```{r}
plot(x=1:5, y=rep(5,5), pch=19, cex=12, col=rgb(.25, .5, .3, alpha=.5), xlim=c(0,6))  
```


```{r}
par(bg="gray40")
col.tr <- grDevices::adjustcolor("557799", alpha=0.7)
plot(x=1:5, y=rep(5,5), pch=19, cex=12, col=col.tr, xlim=c(0,6)) 
```
 
 
```{r}
colors()                          # List all named colors
grep("blue", colors(), value=T)   # Colors that have "blue" in the name
```
 
```{r}
palf <- colorRampPalette(c(rgb(1,1,1, .2),rgb(.8,0,0, .7)), alpha=TRUE)
plot(x=10:1, y=1:10, pch=19, cex=5, col=palf(10)) 
```
 
```{r}
# If you don't have R ColorBrewer already, you will need to install it:
install.packages('RColorBrewer')
library('RColorBrewer')
display.brewer.all()
```
 
```{r}
display.brewer.pal(8, "Set3")
```
 
```{r}
pal3 <- brewer.pal(10, "Set3")
plot(x=10:1, y=10:1, pch=19, cex=6, col=pal3)

plot(x=10:1, y=10:1, pch=19, cex=6, col=rev(pal3)) # backwards
```
 
## 2.2 Fonts in R plots

```{r}
install.packages('extrafont')
library('extrafont')

# Import system fonts - may take a while, so DO NOT run this during the workshop.
font_import() 
fonts() # See what font families are available to you now.
loadfonts(device = "win") # use device = "pdf" for pdf plot output. 
```
 
```{r}
plot(x=10:1, y=10:1, pch=19, cex=3, 
     main="This is a plot", col="orange", 
     family="Arial Black" )
```

## 3 Data format, size, and preparation 
 
```{r}
cur_dir <- getwd()
raw_dir <- "./inst/raw-data/Data files"
setwd(raw_dir)

nodes <- read.csv("Dataset1-Media-Example-NODES.csv", header=T, as.is=T)
links <- read.csv("Dataset1-Media-Example-EDGES.csv", header=T, as.is=T)

setwd(cur_dir)
```
 
### 3.1 Dataset 1: edgelist 
 
```{r}
head(nodes)
head(links)
```
 
 
```{r}
nrow(nodes); length(unique(nodes$id))
nrow(links); nrow(unique(links[,c("from", "to")]))
```
 
```{r}
links <- aggregate(links[,3], links[,-3], sum)
links <- links[order(links$from, links$to),]
colnames(links)[4] <- "weight"
rownames(links) <- NULL
```
 
 
### 3.2 Dataset 2: matrix 
 
```{r}
cur_dir <- getwd()
raw_dir <- "./inst/raw-data/Data files"
setwd(raw_dir)

nodes2 <- read.csv("Dataset2-Media-User-Example-NODES.csv", header=T, as.is=T)
links2 <- read.csv("Dataset2-Media-User-Example-EDGES.csv", header=T, row.names=1)

setwd(cur_dir)
```
 
```{r}
head(nodes2)
head(links2)
```
 
```{r}
links2 <- as.matrix(links2)
dim(links2)
dim(nodes2)
```
 
# 4 Plotting networks with igraph 
 
```{r}
library('igraph')
net <- graph_from_data_frame(d=links, vertices=nodes, directed=T) 
net
```
 
 
```{r}
E(net)       # The edges of the "net" object
V(net)       # The vertices of the "net" object
E(net)$type  # Edge attribute "type"
V(net)$media # Vertex attribute "media"

# Find nodes and edges by attribute:
# (that returns oblects of type vertex sequence/edge sequence)
V(net)[media=="BBC"]
E(net)[type=="mention"]

# You can also examine the network matrix directly:
net[1,]
net[5,7]
```
 
```{r}
# Get an edge list or a matrix:
as_edgelist(net, names=T)
as_adjacency_matrix(net, attr="weight")

# Or data frames describing nodes and edges:
as_data_frame(net, what="edges")
as_data_frame(net, what="vertices")
```
 
```{r}
plot(net) # not a pretty picture!
```
 
```{r}
net <- simplify(net, remove.multiple = F, remove.loops = T) 
```
 
 
```{r}
plot(net, edge.arrow.size=.4,vertex.label=NA)
```
 
```{r}
head(nodes2)
head(links2)

net2 <- graph_from_incidence_matrix(links2)
table(V(net2)$type)
```
 
 
```{r}
# Plot with curved edges (edge.curved=.1) and reduce arrow size:
# Note that using curved edges will allow you to see multiple links
# between two nodes (e.g. links going in either direction, or multiplex links)
plot(net, edge.arrow.size=.4, edge.curved=.1)
```
 
```{r}
# Set edge color to light gray, the node & border color to orange 
# Replace the vertex label with the node names stored in "media"
plot(net, edge.arrow.size=.2, edge.color="orange",
     vertex.color="orange", vertex.frame.color="#ffffff",
     vertex.label=V(net)$media, vertex.label.color="black") 
```
 
```{r}
# Generate colors based on media type:
colrs <- c("gray50", "tomato", "gold")
V(net)$color <- colrs[V(net)$media.type]

# Compute node degrees (#links) and use that to set node size:
deg <- degree(net, mode="all")
V(net)$size <- deg*3
# We could also use the audience size value:
V(net)$size <- V(net)$audience.size*0.6

# The labels are currently node IDs.
# Setting them to NA will render no labels:
V(net)$label <- NA

# Set edge width based on weight:
E(net)$width <- E(net)$weight/6

#change arrow size and edge color:
E(net)$arrow.size <- .2
E(net)$edge.color <- "gray80"
E(net)$width <- 1+E(net)$weight/12
plot(net) 
```
 
```{r}
plot(net, edge.color="orange", vertex.color="gray50") 

```
 
 
### add legend

```{r}
plot(net) 
legend(x=-1.5, y=-1.1, c("Newspaper","Television", "Online News"), pch=21,
       col="#777777", pt.bg=colrs, pt.cex=2, cex=.8, bty="n", ncol=1)
```
 
### plot only labels
 
```{r}
plot(net, vertex.shape="none", vertex.label=V(net)$media, 
     vertex.label.font=2, vertex.label.color="gray40",
     vertex.label.cex=.7, edge.color="gray85")
```
 
Let’s color the edges of the graph based on their source node color. We can get the starting node for each edge with the  ends() igraph function. It returns the start and end vertex for edges listed in the es parameter. The names parameter control whether the function returns edge names or IDs.

```{r}
edge.start <- ends(net, es=E(net), names=F)[,1]
edge.col <- V(net)$color[edge.start]

plot(net, edge.color=edge.col, edge.curved=.1)  

```

## 4.3 Network layouts

```{r}
net.bg <- sample_pa(80) 
V(net.bg)$size <- 8
V(net.bg)$frame.color <- "white"
V(net.bg)$color <- "orange"
V(net.bg)$label <- "" 
E(net.bg)$arrow.mode <- 0
plot(net.bg)
```

```{r}
plot(net.bg, layout=layout_randomly)
```
 
```{r}
l <- layout_in_circle(net.bg)
plot(net.bg, layout=l)
```
 
```{r}
l <- cbind(1:vcount(net.bg), c(1, vcount(net.bg):2))
plot(net.bg, layout=l)
```
 
```{r}
# Randomly placed vertices
l <- layout_randomly(net.bg)
plot(net.bg, layout=l)

```
 
```{r}
# Circle layout
l <- layout_in_circle(net.bg)
plot(net.bg, layout=l)
```
 
```{r}
# 3D sphere layout
l <- layout_on_sphere(net.bg)
plot(net.bg, layout=l)
```
 
### Fruchterman-Reingold 

```{r}
l <- layout_with_fr(net.bg)
plot(net.bg, layout=l)
```
 
```{r}
par(mfrow=c(2,2), mar=c(0,0,0,0))   # plot four figures - 2 rows, 2 columns
plot(net.bg, layout=layout_with_fr)
plot(net.bg, layout=layout_with_fr)
plot(net.bg, layout=l)
plot(net.bg, layout=l)
```
 
```{r}
dev.off()
```
 
```{r}
l <- layout_with_fr(net.bg)
l <- norm_coords(l, ymin=-1, ymax=1, xmin=-1, xmax=1)

par(mfrow=c(2,2), mar=c(0,0,0,0))
plot(net.bg, rescale=F, layout=l*0.4)
plot(net.bg, rescale=F, layout=l*0.6)
plot(net.bg, rescale=F, layout=l*0.8)
plot(net.bg, rescale=F, layout=l*1.0)
```
 
```{r}
dev.off()
```
 
### Kamada-Kawai
 
```{r}
l <- layout_with_kk(net.bg)
plot(net.bg, layout=l)
```
 
### LGL algorithm

```{r}
plot(net.bg, layout=layout_with_lgl)
```
 
### Multidimensional algorithm

```{r}
plot(net.bg, layout=layout_with_mds)
```
 
### available layouts

```{r}
layouts <- grep("^layout_", ls("package:igraph"), value=TRUE)[-1] 
# Remove layouts that do not apply to our graph.
layouts <- layouts[!grepl("bipartite|merge|norm|sugiyama|tree", layouts)]

par(mfrow=c(3,3), mar=c(1,1,1,1))
for (layout in layouts) {
  print(layout)
  l <- do.call(layout, list(net)) 
  plot(net, edge.arrow.mode=0, layout=l, main=layout) }
```
 
## 4.4 Highlighting aspects of the network

```{r}
hist(links$weight)
mean(links$weight)
sd(links$weight)
```

```{r}
cut.off <- mean(links$weight) 
net.sp <- delete_edges(net, E(net)[weight<cut.off])
plot(net.sp) 
```

```{r}
par(mfrow=c(1,2))

# Community detection based on label propagation:
clp <- cluster_label_prop(net)
class(clp)

# Community detection returns an object of class "communities" 
# which igraph knows how to plot: 
plot(clp, net)
 
# We can also plot the communities without relying on their built-in plot:
V(net)$community <- clp$membership
colrs <- adjustcolor( c("gray50", "tomato", "gold", "yellowgreen"), alpha=.6)
plot(net, vertex.color=colrs[V(net)$community])

dev.off()
```

## 4.5 Highlighting specific nodes or links

```{r}
dist.from.NYT <- distances(net, v=V(net)[media=="NY Times"], 
                           to=V(net), weights=NA)

# Set colors to plot the distances:
oranges <- colorRampPalette(c("dark red", "gold"))
col <- oranges(max(dist.from.NYT)+1)
col <- col[dist.from.NYT+1]

plot(net, vertex.color=col, vertex.label=dist.from.NYT, edge.arrow.size=.6, 
     vertex.label.color="white")
```

```{r}
news.path <- shortest_paths(net, 
                            from = V(net)[media=="MSNBC"], 
                             to  = V(net)[media=="New York Post"],
                             output = "both") # both path nodes and edges

# Generate edge color variable to plot the path:
ecol <- rep("gray80", ecount(net))
ecol[unlist(news.path$epath)] <- "orange"
# Generate edge width variable to plot the path:
ew <- rep(2, ecount(net))
ew[unlist(news.path$epath)] <- 4
# Generate node color variable to plot the path:
vcol <- rep("gray40", vcount(net))
vcol[unlist(news.path$vpath)] <- "gold"

plot(net, vertex.color=vcol, edge.color=ecol, 
     edge.width=ew, edge.arrow.mode=0)
```

```{r}
inc.edges <- incident(net,  V(net)[media=="Wall Street Journal"], mode="all")

# Set colors to plot the selected edges.
ecol <- rep("gray80", ecount(net))
ecol[inc.edges] <- "orange"
vcol <- rep("grey40", vcount(net))
vcol[V(net)$media=="Wall Street Journal"] <- "gold"
plot(net, vertex.color=vcol, edge.color=ecol)
```

```{r}
neigh.nodes <- neighbors(net, V(net)[media=="Wall Street Journal"], mode="out")

# Set colors to plot the neighbors:
vcol[neigh.nodes] <- "#ff9d00"
plot(net, vertex.color=vcol)
```

```{r}
par(mfrow=c(1,2))
plot(net, mark.groups=c(1,4,5,8), mark.col="#C5E5E7", mark.border=NA)

# Mark multiple groups:
plot(net, mark.groups=list(c(1,4,5,8), c(15:17)), 
          mark.col=c("#C5E5E7","#ECD89A"), mark.border=NA)

dev.off()
```

## 4.6 Interactive plotting with tkplot

```{r}
tkid <- tkplot(net) #tkid is the id of the tkplot that will open
l <- tkplot.getcoords(tkid) # grab the coordinates from tkplot
plot(net, layout=l)
```

### quick heatmap 

```{r}
netm <- as_adjacency_matrix(net, attr="weight", sparse=F)
colnames(netm) <- V(net)$media
rownames(netm) <- V(net)$media

palf <- colorRampPalette(c("gold", "dark orange")) 
heatmap(netm[,17:1], Rowv = NA, Colv = NA, col = palf(100), 
        scale="none", margins=c(10,10) )
```

### simple graphs

```{r}
# Plot the egree distribution for our network:
deg.dist <- degree_distribution(net, cumulative=T, mode="all")
plot( x=0:max(deg), y=1-deg.dist, pch=19, cex=1.2, col="orange", 
      xlab="Degree", ylab="Cumulative Frequency")

```

## 4.8 Plotting two-mode networks


```{r}
head(nodes2)
head(links2)

net2
plot(net2, vertex.label=NA)
```

```{r}
# Media outlets are blue squares, audience nodes are orange circles:
V(net2)$color <- c("steel blue", "orange")[V(net2)$type+1]
V(net2)$shape <- c("square", "circle")[V(net2)$type+1]

# Media outlets will have name labels, audience members will not:
V(net2)$label <- ""
V(net2)$label[V(net2)$type==F] <- nodes2$media[V(net2)$type==F] 
V(net2)$label.cex=.6
V(net2)$label.font=2

plot(net2, vertex.label.color="white", vertex.size=(2-V(net2)$type)*8) 
```

```{r}
plot(net2, vertex.label=NA, vertex.size=7, layout=layout.bipartite) 
```

### text as nodes

```{r}
plot(net2, vertex.shape="none", vertex.label=nodes2$media,
     vertex.label.color=V(net2)$color, vertex.label.font=2, 
     vertex.label.cex=.6, edge.color="gray70",  edge.width=2)
```

### images as nodes

```{r}
# install.packages('png')
library('png')
 
img.1 <- readPNG("./inst/raw-data/Data files/images/news.png")
img.2 <- readPNG("./inst/raw-data/Data files/images/user.png")

V(net2)$raster <- list(img.1, img.2)[V(net2)$type+1]

plot(net2, vertex.shape="raster", vertex.label=NA,
     vertex.size=16, vertex.size2=16, edge.width=2)
```

```{r}
plot(net2, vertex.shape="raster", vertex.label=NA,
     vertex.size=16, vertex.size2=16, edge.width=2)

img.3 <- readPNG("./inst/raw-data/Data files/images/puppy.png")
rasterImage(img.3,  xleft=-1.7, xright=0, ybottom=-1.2, ytop=0)
```

### bipartite projections for the two-mode network

```{r}
par(mfrow=c(1,2))

net2.bp <- bipartite.projection(net2) 

plot(net2.bp$proj1, vertex.label.color="black", vertex.label.dist=1,
     vertex.label=nodes2$media[!is.na(nodes2$media.type)])

plot(net2.bp$proj2, vertex.label.color="black", vertex.label.dist=1,
     vertex.label=nodes2$media[ is.na(nodes2$media.type)]) 
```


## 4.9 Plotting multiplex networks

```{r}
E(net)$width <- 1.5
plot(net, edge.color=c("dark red", "slategrey")[(E(net)$type=="hyperlink")+1],
      vertex.color="gray40", layout=layout_in_circle, edge.curved=.3)
```

```{r}
net.m <- net - E(net)[E(net)$type=="hyperlink"] # another way to delete edges:
net.h <- net - E(net)[E(net)$type=="mention"]   # using the minus operator

# Plot the two links separately:
par(mfrow=c(1,2))
plot(net.h, vertex.color="orange", main="Tie: Hyperlink")
plot(net.m, vertex.color="lightsteelblue2", main="Tie: Mention")
```

```{r}
# Make sure the nodes stay in place in both plots:
l <- layout_with_fr(net)
plot(net.h, vertex.color="orange", layout=l, main="Tie: Hyperlink")
plot(net.m, vertex.color="lightsteelblue2", layout=l, main="Tie: Mention")

# dev.off()
```

###  two nodes and three ties between

```{r}
multigtr <- graph( edges=c(1,2, 1,2, 1,2), n=2 )
l <- layout_with_kk(multigtr)

# Let's just plot the graph:
plot(multigtr, vertex.color="lightsteelblue", vertex.frame.color="white",
     vertex.size=40, vertex.shape="circle", vertex.label=NA,
     edge.color=c("gold", "tomato", "yellowgreen"), edge.width=5,
     edge.arrow.size=3, edge.curved=0.1, layout=l)
```

```{r}
plot(multigtr, vertex.color="lightsteelblue", vertex.frame.color="white", 
     vertex.size=40, vertex.shape="circle", vertex.label=NA,
     edge.color=c("gold", "tomato", "yellowgreen"), edge.width=5,
     edge.arrow.size=3, edge.curved=curve_multiple(multigtr), layout=l)

```


```{r}
detach('package:igraph')
```



# 5 Quick example using the network package

```{r}
library('network')

net3 <- network(links,  vertex.attr=nodes, matrix.type="edgelist", 
                loops=F, multiple=F, ignore.eval = F)
```

```{r}
net3[,]
net3 %n% "net.name" <- "Media Network" #  network attribute
net3 %v% "media"    # Node attribute
net3 %e% "type"     # Node attribute
```

```{r}
net3 %v% "col" <- c("gray70", "tomato", "gold")[net3 %v% "media.type"]
plot(net3, vertex.cex=(net3 %v% "audience.size")/7, vertex.col="col")
```

```{r}
l <- plot(net3, vertex.cex=(net3 %v% "audience.size")/7, vertex.col="col")
plot(net3, vertex.cex=(net3 %v% "audience.size")/7, vertex.col="col", coord=l)
```





```{r}
library('network')
plot(net3, vertex.cex=(net3 %v% "audience.size")/7, vertex.col="col", interactive=T)
```

```{r}
detach('package:network')
```


## 6.2 Interactive JS visualization with visNetwork

```{r}
library('visNetwork') 
visNetwork(nodes, links, width="100%", height="400px", main="Network!")
```


```{r}
nodes$shape <- "dot"  
nodes$shadow <- TRUE # Nodes will drop shadow
nodes$title <- nodes$media # Text on click
nodes$label <- nodes$type.label # Node label
nodes$size <- nodes$audience.size # Node size
nodes$borderWidth <- 2 # Node border width

nodes$color.background <- c("slategrey", "tomato", "gold")[nodes$media.type]
nodes$color.border <- "black"
nodes$color.highlight.background <- "orange"
nodes$color.highlight.border <- "darkred"

visNetwork(nodes, links)
```

```{r}
links$width <- 1+links$weight/8 # line width
links$color <- "gray"    # line color  
links$arrows <- "middle" # arrows: 'from', 'to', or 'middle'
links$smooth <- FALSE    # should the edges be curved?
links$shadow <- FALSE    # edge shadow

visNetwork(nodes, links)
```

```{r}
visNetwork(nodes, links) %>%
  visOptions(highlightNearest = TRUE, 
             selectedBy = "type.label")
```

## 6.3 Interactive JS visualization with threejs


```{r}
library(threejs)
library(htmlwidgets)
library(igraph)

graphjs(net) 
```


```{r}
net.js <- igraph2graphjs(net)
net.js$nodes[,2] <- NULL #delete duplicated label column
net.js$nodes$label <- net.js$nodes$media
net.js$nodes$size <- net.js$nodes$size/10
```

```{r}
graphjs(net.js, main="Network!", bg="gray10", showLabels=F, stroke=F, 
        curvature=0.1, attraction=0.9, repulsion=0.8, opacity=0.9)
```

## 6.4 Interactive JS visualization with networkD3

```{r}
library(networkD3)

el <- data.frame(from=as.numeric(factor(links$from))-1, 
                 to=as.numeric(factor(links$to))-1 )
```

```{r}
nl <- cbind(idn=factor(nodes$media, levels=nodes$media), nodes) 

```



```{r}
forceNetwork(Links = el, Nodes = nl, Source="from", Target="to",
               NodeID = "idn", Group = "type.label",linkWidth = 1,
               linkColour = "#afafaf", fontSize=12, zoom=T, legend=T,
               Nodesize=6, opacity = 0.8, charge=-300, 
               width = 600, height = 400)
```



# 8 Overlaying networks on geographic maps

```{r}
install.packages('maps')
install.packages('geosphere')

library('maps')
library('geosphere')

```



```{r}
par(mfrow = c(2,2), mar=c(0,0,0,0))

map("usa", col="tomato",  border="gray10", fill=TRUE, bg="gray30")
map("state", col="orange",  border="gray10", fill=TRUE, bg="gray30")
map("county", col="palegreen",  border="gray10", fill=TRUE, bg="gray30")
map("world", col="skyblue",  border="gray10", fill=TRUE, bg="gray30")
```


```{r}
airports <- read.csv("./inst/raw-data/Data files/Dataset3-Airlines-NODES.csv", header=TRUE) 
flights  <- read.csv("./inst/raw-data/Data files/Dataset3-Airlines-EDGES.csv", header=TRUE, as.is=TRUE)

head(flights)
```

```{r}
head(airports)
```

```{r}
# Select only large airports: ones with more than 10 connections in the data.
tab <- table(flights$Source)
big.id <- names(tab)[tab>10]

airports <- airports[airports$ID %in% big.id,]
flights  <- flights[flights$Source %in% big.id & 
                    flights$Target %in% big.id, ]
```


```{r}


```


For each flight in our data, we will use `gcIntermediate()` to generate the coordinates of the shortest arc that connects its start and end point (think distance on the surface of a sphere). After that, we will plot each arc over the map using lines().


```{r}
# Plot a map of the united states:
map("state", col="grey20", fill=TRUE, bg="black", lwd=0.1)

# Add a point on the map for each airport:
points(x=airports$longitude, y=airports$latitude, pch=19, 
       cex=airports$Visits/80, col="orange")

col.1 <- adjustcolor("orange red", alpha=0.4)
col.2 <- adjustcolor("orange", alpha=0.4)
edge.pal <- colorRampPalette(c(col.1, col.2), alpha = TRUE)
edge.col <- edge.pal(100)

for(i in 1:nrow(flights))  {
    node1 <- airports[airports$ID == flights[i,]$Source,]
    node2 <- airports[airports$ID == flights[i,]$Target,]
    
    arc <- gcIntermediate( c(node1[1,]$longitude, node1[1,]$latitude), 
                           c(node2[1,]$longitude, node2[1,]$latitude), 
                           n=1000, addStartEnd=TRUE )
    
    edge.ind <- round(100*flights[i,]$Freq / max(flights$Freq))
    
    lines(arc, col=edge.col[edge.ind], lwd=edge.ind/30)
}
```

